Dependent Object Types

### 2012-09-21 doc/fool/article.tex
f-bounded quantification ex

\subsection{Abstract Type Members}
The \code{choices.Alt} trait takes three abstract type members:
\code{C}, \code{A}, \code{B}. \code{A} and \code{B} are upper bounded
by \code{C}. The intention is that the \code{choose} function takes an
\code{A} and a \code{B} and returns one or the other.
\begin{lstlisting}
object choices {
  trait Alt {
    type C
    type A <: C
    type B <: C
    val choose : A => B => C
  }
}
\end{lstlisting}
In DOT, we can state more precisely the return type of $\mi{choose}$,
thanks to union types:
\begin{align*}
&\new {\mi{choices}} {\Top \mlrefine z {\\
&\gap\Ldecl {\mi{Alt_c}} {\Bot} {\Top \mlrefine a {\\
&\gap\gap\Ldecl C \Bot \Top\\
&\gap\gap\Ldecl A \Bot {a.C}\\
&\gap\gap\Ldecl B \Bot {a.C}\\
&\gap\gap\mdecl {\mi{choose}} {a.A} {\arrow {a.B} {a.A \tor a.B}}\\
&\gap}}\\
&}\ldefs{}}
\end{align*}

@@

>>>>

\subsection{Abstract Type Members}
The \code{choices.Alt} trait takes three abstract type members:
\code{C}, \code{A}, \code{B}. \code{A} and \code{B} are upper bounded
by \code{C}. The intention is that the \code{choose} function takes an
\code{A} and a \code{B} and returns one or the other.
\begin{lstlisting}
object choices {
  trait Alt {
    type C
    type A <: C
    type B <: C
    val choose : A => B => C
  }
}
\end{lstlisting}
In DOT, we can state more precisely the return type of $\mi{choose}$,
thanks to union types:
\begin{align*}
&\new {\mi{choices}} {\Top \mlrefine z {\\
&\gap\Ldecl {\mi{Alt_c}} {\Bot} {\Top \mlrefine a {\\
&\gap\gap\Ldecl C \Bot \Top\\
&\gap\gap\Ldecl A \Bot {a.C}\\
&\gap\gap\Ldecl B \Bot {a.C}\\
&\gap\gap\mdecl {\mi{choose}} {a.A} {\arrow {a.B} {a.A \tor a.B}}\\
&\gap}}\\
&}\ldefs{}}{}
\end{align*}

@@

\subsection{F-bounded Quantification}

F-bounded quantification describes an upper bound that itself contains
the type being constrained: for example, \code{Int extends Ordered[Int]}.
Here, we define \code{MetaAlt} to extend
\code{choices.Alt} with \code{C} as an alias for \code{MetaAlt}.

\begin{lstlisting}
trait MetaAlt extends choices.Alt {
  type C = MetaAlt
  type A = C
  type B = C
}
\end{lstlisting}

Now, we can define some \code{MetaAlt} instances:
\begin{lstlisting}
val first = new MetaAlt {
  val choose: C => C => C = a => b => a
}
val last = new MetaAlt {
  val choose: C => C => C = a => b => b
}
val recfirst = new MetaAlt {
  val choose: C => C => C = a => b => a.choose(a)(b)
}
val reclast = new MetaAlt {
  val choose: C => C => C = a => b => b.choose(a)(b)
}
\end{lstlisting}

The equivalent in DOT is straightforward. We wrap $\mi{MetaAlt}$ in a
namespace, so that we can refer to it.
\begin{align*}
&\mlnew m {\Top \mlrefine m {\\
&\gap\Ldecl {\mi{MetaAlt_c}} {\Bot} {{\mi{choices.Alt_c}} \mlrefine a {\\
&\gap\gap\Ldecl C {\mi{m.MetaAlt_c}} {\mi{m.MetaAlt_c}}\\
&\gap\gap\Ldecl A {\mi{a.C}} {\mi{a.C}}\\
&\gap\gap\Ldecl B {\mi{a.C}} {\mi{a.C}}\\
&\gap}}\\
&}\ldefs{}}{}
\end{align*}

Now, we can create the equivalent of \code{first} ($f$), \code{last}
($l$), \code{recfirst} ($\mi{rf}$) and \code{reclast} ($\mi{rl}$):
\begin{align*}
&\mlnew f {{\mi{m.MetaAlt_c}}\mlldefs{\\
&\gap\gap\mi{choose}(a)=\fun b {\mi{m.MetaAlt_c}} {\mi{m.MetaAlt_c}} a}}{
\mlnew l {{\mi{m.MetaAlt_c}}\mlldefs{\\
&\gap\gap\mi{choose}(a)=\fun b {\mi{m.MetaAlt_c}} {\mi{m.MetaAlt_c}} b}}{
\mlnew {\mi{rf}} {{\mi{m.MetaAlt_c}}\mlldefs{\\
&\gap\gap\mi{choose}(a)=\fun b {\mi{m.MetaAlt_c}} {\mi{m.MetaAlt_c}} {\\
&\gap\gap\gap \app {{\mi{a.choose}}(a)} b}}}{
\mlnew {\mi{rf}} {{\mi{m.MetaAlt_c}}\mlldefs{\\
&\gap\gap\mi{choose}(a)=\fun b {\mi{m.MetaAlt_c}} {\mi{m.MetaAlt_c}} {\\
&\gap\gap\gap \app {{\mi{b.choose}}(a)} b}}}{
}}}}
\end{align*}

Given these definitions, here is a valid expression, which evaluates to $f$:
$\cast \Top \app {{\mi{rf.choose}}(f)} l$.
